Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-09-02 paper.tex
Fix 3 images
Fix 2 images
Fix formatting

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle and
arbitrary triangle pattern. The former was believed to be faster to compute, but
with the help of precomputed regular positions we achieved the same speed for
arbitrary pattern. In this work we also implement and demonstrate computations
of directional 3-point CFs -- for this purpose right-triangular pattern seems to
be superior due to explicit orientation and high coverage.  Moreover, we assess
the errors in CFs' evaluation due to image or pattern rotations and show that
they have minor effect on accuracy of computations.  The execution times of our
algorithms for the same number of samples are orders of magnitude lower than in
existing published counterparts. We showed that the volume of data produced get
unwieldy very easily, especially if computations are performed in frequency
domain. For these reasons until information content of different sets of
correlation functions with different "n-pointness" is not known, advantages of
CFs with $n>3$ are not clear. Nonetheless, developed algorithms and code are
universal enough to be easily extendable to any $n$ (with increasing
computational and RAM burden). All results are available as part of open source
package \code{CorrelationFunctions.jl} [Postnicov et al., 2024,
  Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence,zubov2024search}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation. In fact, we need a single universal descriptor defined by only
one characteristic -- information content. The latter is an ability of the
metric to describe the structure at hand with 100\% meaning the ability to
fully recover the structure back from this metric. Thus, universal descriptor
has to possess following key properties: (1) high information content, (2) ease
of assess, (3) high compression ratio (i.e., descriptor is much "smaller" than
the structure itself).

@@

  \includegraphics[width=0.8\linewidth]{images/3point.png}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/web-like.png}
  \caption[]{Schematic representation of patterns for 3-point correlation
    functions evaluation: arbitrary or random triangle, weblike, and
    right-triangular patterns. Note that real patterns are more complex and are
    made simple for visualization, i.e., for right-triangular pattern we sample
    all possible variations between points. Equilateral or isosceles triangles
    can be considered as arbitrary patterns and can be either fixed in space, or
    rotated randomly.}
  \label{fig:pattern}
\end{figure*}

@@

\section{Summary}
In this paper, we have considered the problem of computing n-point correlation
functions, namely $S_n, C_n, F_{s...s}$ and $F_{s...v}$, for which
computationally efficient algorithms can be developed (including computations on
GPU). For $n=3$ we have developed code in Julia language to compute $S_3, C_3,
F_{sss}, F_{ssv}$ and $F_{svv}$ in an arbitrary set of points. We also provide a
function which generates a right-angle triangle pattern which can be useful for
directional CFs evaluation. Computations are easily
vectorized and utilize the power of modern CPU and GPU architectures. The
execution times for the same number of sample are orders of magnitude lower than
existing published counterparts. We showed that the volume of data produced get
unwieldy very easily, especially if computations are performed in frequency
domain. For these reasons until information content of different sets of
correlation functions with different "n-pointness" is known, it is seemingly no
much reason to increase the number of points. Nonetheless, developed algorithms
and code are universal enough to be easily extendable to any $n$ (with
increasing computational and RAM burden). All results are available as part of
open source package \code{CorrelationFunctions.jl}, developed by our group and
freely available to the public with numerous Jupiter notebooks showing examples
of applications to assess 3-point correlation functions for 2D and 3D images.

>>>>

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle and
arbitrary triangle pattern. The former was believed to be faster to compute, but
with the help of precomputed regular positions we achieved the same speed for
arbitrary pattern. In this work we also implement and demonstrate computations
of directional 3-point CFs -- for this purpose right-triangular pattern seems to
be superior due to explicit orientation and high coverage.  Moreover, we assess
the errors in CFs' evaluation due to image or pattern rotations and show that
they have minor effect on accuracy of computations.  The execution times of our
algorithms for the same number of samples are orders of magnitude lower than in
existing published counterparts. We showed that the volume of data produced get
unwieldy very easily, especially if computations are performed in frequency
domain. For these reasons until information content of different sets of
correlation functions with different ``n-pointness'' is not known, advantages of
CFs with $n>3$ are not clear. Nonetheless, developed algorithms and code are
universal enough to be easily extendable to any $n$ (with increasing
computational and RAM burden). All results are available as part of open source
package \code{CorrelationFunctions.jl} [Postnicov et al., 2024,
  Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence,zubov2024search}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation. In fact, we need a single universal descriptor defined by only
one characteristic -- information content. The latter is an ability of the
metric to describe the structure at hand with 100\% meaning the ability to
fully recover the structure back from this metric. Thus, universal descriptor
has to possess following key properties: (1) high information content, (2) ease
of assess, (3) high compression ratio (i.e., descriptor is much ``smaller'' than
the structure itself).

@@

  \subfigure[2-point CF]{
    \includegraphics[width=0.45\linewidth]{images/2p.png}}
  \hfill
  \subfigure[3-point CF]{
    \includegraphics[width=0.45\linewidth]{images/3p.png}}

@@

\begin{figure*}[ht]
  \centering
  \subfigure[Arbitrary]{
    \includegraphics[width=0.3\linewidth]{images/arbitrary.png}}
  \hfill
  \subfigure[Weblike]{
    \includegraphics[width=0.3\linewidth]{images/web.png}}
  \hfill
  \subfigure[Right-triangular]{
    \includegraphics[width=0.3\linewidth]{images/right.png}}
  \caption[]{Schematic representation of patterns for 3-point correlation
    functions evaluation: arbitrary or random triangle, weblike, and
    right-triangular patterns. Note that real patterns are more complex and are
    made simple for visualization, i.e., for right-triangular pattern we sample
    all possible variations between points. Equilateral or isosceles triangles
    can be considered as arbitrary patterns and can be either fixed in space, or
    rotated randomly.}
  \label{fig:pattern}
\end{figure*}

@@

\section{Summary}
In this paper, we have considered the problem of computing n-point correlation
functions, namely $S_n, C_n, F_{s...s}$ and $F_{s...v}$, for which
computationally efficient algorithms can be developed (including computations on
GPU). For $n=3$ we have developed code in Julia language to compute $S_3, C_3,
F_{sss}, F_{ssv}$ and $F_{svv}$ in an arbitrary set of points. We also provide a
function which generates a right-angle triangle pattern which can be useful for
directional CFs evaluation. Computations are easily
vectorized and utilize the power of modern CPU and GPU architectures. The
execution times for the same number of sample are orders of magnitude lower than
existing published counterparts. We showed that the volume of data produced get
unwieldy very easily, especially if computations are performed in frequency
domain. For these reasons until information content of different sets of
correlation functions with different ``n-pointness'' is known, it is seemingly no
much reason to increase the number of points. Nonetheless, developed algorithms
and code are universal enough to be easily extendable to any $n$ (with
increasing computational and RAM burden). All results are available as part of
open source package \code{CorrelationFunctions.jl}, developed by our group and
freely available to the public with numerous Jupiter notebooks showing examples
of applications to assess 3-point correlation functions for 2D and 3D images.
